---
title: "Prediction Assignment Writeup - ML "
author: "Colleen"
date: "4/13/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Assignment Introduction

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.

# Assignment Deliverables

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

# Sources

Information is available at: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har

Training Data:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

Test Data: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

Data for the Project:
http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har.

Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.


# Loading and Cleaning My Data

Required R Packages
```{r warning=FALSE}
library(lattice)
library(ggplot2)
library(caret)
library(rpart)
library(rpart.plot)
library(corrplot)
library(rattle)
library(randomForest)
library(RColorBrewer)

set.seed(222)
```

Load data for training and test sets

```{r}
url_train <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
url_quiz  <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

data_train <- read.csv(url(url_train), strip.white = TRUE, na.strings = c("NA",""))
data_quiz  <- read.csv(url(url_quiz),  strip.white = TRUE, na.strings = c("NA",""))

dim(data_train)
```

```{r}
dim(data_quiz)
```

Create 2 partitions (75% & 25%) within training set

```{r}
in_train  <- createDataPartition(data_train$classe, p=0.75, list=FALSE)
train_set <- data_train[ in_train, ]
test_set  <- data_train[-in_train, ]

dim(train_set)
```

```{r}
dim(test_set)
```

Remove NA values and near-zero variance variables, both to be removed together. 

```{r}
nzv_var <- nearZeroVar(train_set)

train_set <- train_set[ , -nzv_var]
test_set  <- test_set [ , -nzv_var]

dim(train_set)
```

```{r}
dim(test_set)
```

Remove variables that are mostly NA, a threshold of 95% is selected. 

```{r}
na_var <- sapply(train_set, function(x) mean(is.na(x))) > 0.95
train_set <- train_set[ , na_var == FALSE]
test_set  <- test_set [ , na_var == FALSE]

dim(train_set)
```

```{r}
dim(test_set)
```

Simce columns 1 to 5 are identification variables only, they will be removed as well. 

```{r}
train_set <- train_set[ , -(1:5)]
test_set  <- test_set [ , -(1:5)]

dim(train_set)
```

```{r}
dim(test_set)
```

The number of variables has been reduced from 160 to 54 through cleaning the data. 

# Correlation Analysis

```{r}
corr_matrix <- cor(train_set[ , -54])
corrplot(corr_matrix, order = "FPC", method = "circle", type = "lower",
         tl.cex = 0.6, tl.col = rgb(0, 0, 0))
```
The darker shade of each of the color shows the correlations; the darker blue showing a positive correlation and the darker red showing a negative correlation. Due to so few strong correlations, a few prediction models will be built for better accuracy. 

# Prediction Models

## Decision Tree Model

```{r}
set.seed(2222)
fit_decision_tree <- rpart(classe ~ ., data = train_set, method="class")
fancyRpartPlot(fit_decision_tree)
```

Predictions of the decision tree model with test_set

```{r}
predict_decision_tree <- predict(fit_decision_tree, newdata = test_set, type="class")
conf_matrix_decision_tree <- confusionMatrix(predict_decision_tree, factor(test_set$classe))
conf_matrix_decision_tree
```

The predictive accuracy of the decision tree model is relatively low at 75.2 %.

Plot the predictive accuracy of the decision tree model.

```{r}
plot(conf_matrix_decision_tree$table, col = conf_matrix_decision_tree$byClass, 
     main = paste("Decision Tree Model: Predictive Accuracy =",
                  round(conf_matrix_decision_tree$overall['Accuracy'], 4)))
```

# Generalized Boosted Model (GBM)

```{r}
set.seed(2222)
ctrl_GBM <- trainControl(method = "repeatedcv", number = 5, repeats = 2)
fit_GBM  <- train(classe ~ ., data = train_set, method = "gbm",
                  trControl = ctrl_GBM, verbose = FALSE)
fit_GBM$finalModel
```

Predictions of the GBM on test_set

```{r}
predict_GBM <- predict(fit_GBM, newdata = test_set)
conf_matrix_GBM <- confusionMatrix(predict_GBM, factor(test_set$classe))
conf_matrix_GBM
```
The predictive accuracy of GBM is 98.57%

# Random Forest Model

```{r}
set.seed(2222)
ctrl_RF <- trainControl(method = "repeatedcv", number = 5, repeats = 2)
fit_RF  <- train(classe ~ ., data = train_set, method = "rf",
                  trControl = ctrl_RF, verbose = FALSE)
fit_RF$finalModel
```

Predictions of the Random Forest model on test_set

```{r}
predict_RF <- predict(fit_RF, newdata = test_set)
conf_matrix_RF <- confusionMatrix(predict_RF, factor(test_set$classe))
conf_matrix_RF
```

Predictive accuracy of the Random Forest model is excellent = 99.8%

# Applying the Best Predictive Model to the Test Data

Predictive accuracy of the three models:

-Decision Tree Model: 75.20%

-Generalized Boosted Model: 98.57%

-Random Forest Model: 99.80%

The Random Forest Model is selected and used to make predictions on the 20 data points from the original testing dataset (data_quiz)

```{r}
predict_quiz <- as.data.frame(predict(fit_RF, newdata = data_quiz))
predict_quiz
```



